\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-10-04 chapters/bci.tex
Update bttda formulation

\todo{comments liesel paradigms}
\todo{spellcheck}
\todo{are all acronyms defined}
\todo{fix all latex warnings to make sure no refs are missing (or grep on that
specific warning)}
\todo{uniform ref or autoref}
\todo{don't capitalize figure, table, algorithm,...}
\todo{define short captions for tables and figures}
\todo{fix lacheck/chktex notifications}
\todo{use terminology individuals with severe
speech and physical impairments (SSPI), avoid phrasing 'suffering from'}
\todohsm{more subtitles}
\todofc{double check statements about what has been published 'for the sake of clarity, the introduction has been mofed to"}
\todofc{chapters should look more like full papers}
\todo{uniform vsa condition vsa setting}
\todo{acronyms written in full every new chapter and in captions}
\todo{search for ??}

@@

\subsection{Active \& passive \ac{bci}}

@@

\subsection{Reactive \ac{bci}}
Reactive \ac{bci} take another approach, by providing a discrete set of
sensory stimulations towards which the user's attention can directed.
Compared to active \ac{bci}, reactive paradigms can more easily induce fatigue
due to the constant stimulation.
These paradigms are also somewhat less intuitive compared to e.g.\ speech, and
their expressive power is limited by how many different stimuli can effectively
presented and attended within a reasonable timeframe.
Nevertheless, reactive \ac{bci} work well with \ac{eeg} recordings, and, most
importantly, they work for most people~\cite{Allison2010a,Edlinger2014}.

@@

The epochs contain values representing the voltage of a specific channel relative to the
reference at a specific time relative to the stimulus.
In \ac{erp} classification, they are usually directly used as features, by
concatenating all channels in the matrix format epoch $X$ into one feature
vector s.t.\ $\mat{x} = \mat\vec(X)$.

\subsection{Decoders}
Several techniques are suited to decode~\ac{erps}.
Below are some methods that can achieve state-of-the art performance and are
relevant to this work.

\subsubsection{\Acs{lcmv} beamforming}
Linear decoding techniques such as the \ac{lcmv}
beamformer~\cite{Wittevrongel2016} generally work well for
classifying \ac{erp} epochs.
\Ac{lcmv} beamforming learns a weight vector $\mat{w}$ from a set of epochs that can
be multiplied with the feature vector of a new epoch to obtain a prediction
score.
The weights $\mat{w}$ are obtained by minimizing the variance of the output
\begin{equation}
  \argmin_\mat{w}\mat{w}^\intercal\mat{C}
	\mat{w}^\intercal
\end{equation}
under the linear constraint
\begin{equation}
	\mat{a}^\intercal\mat{w} = 1
\end{equation}
with activation pattern $a$ the difference between the average of the attended epochs and the
average of the non-attended epochs in the training data.
This is then solved by  method of Lagrange multipliers as
\begin{equation}
	\mat{w} =
  \frac{\mat{C}^{-1}\mat{a}^\intercal}
  {\mat{a}\mat{C}^{-1}\mat{a}^\intercal}
\end{equation}
with $\mat{C}$ the background covariance.

\subsubsection{\Ac{lda}}
\Ac{lda} is a more well-known method that is, int fact, nearly equivalent to
\ac{lcmv} beamforming~\cite{Treder2016} under certain conditions.
It assumes normally distributed data with a different mean per class but a
common covariance.
Similar to \ac{lcmv}-beamforming, \ac{lda} learns a weight vector $\mat{w}$
that satisfies
\begin{equation}
  \argmax_\mat{w}\frac{\mat{w}^\intercal\mat{S}_\text{b}\mat{w}}{\mat{w}^\intercal\mat{S}_\text{w}\mat{w}}
\end{equation}
The between-class scatter matrix $\mat{S}_\text{b}$ models the variability
between the expected responses, the within-class scatter matrix
$\mat{S}_\text{w}$ the expected noise.
The optimization criterion corresponds to finding a projection that maximizes
the spread between classes (signal) while minimizing the spread within classes
(noise).
$mat{w}$ can be obtained as eigenvector of
$\mat{S}_\text{w}^{-1}\mat{S}_\text{b}$ corresponding to the largest eigenvalue.

@@

\subsubsection{\Ac{tlda}}

Another example of this is \ac{tlda}~\cite{Sosulski2022}, which assumes temporal
stationarity on the background noise to regularize the between-class covariance
matrix of \ac{lda}
\begin{equation}
  \mat{S}_\text{w} = \begin{bmatrix}
s_0 & s_1 & s_2 & \cdots & s_{n-1} \\
s_1 & s_0 & s_1 & \cdots & s_{n-2} \\
s_2 & s_1 & s_0 & \cdots & s_{n-3} \\
\vdots & \vdots & \vdots & \ddots & \vdots \\
s_{n-1} & s_{n-2} & s_{n-3} & \cdots & s_0
\end{bmatrix}
\end{equation}
This method is simple yet effective reaching state-of-the-art decoding
performance.

@@

With this distance defined, the epochs can than be either directly be
classified on the Riemannian manifold using a minimum distance to mean
classifier, or they can be projected to the tangent space.
This is a euclidean space that locally approximates the Riemannian distance,
which allows us to apply regular machine learning methods like an SVM or
\ac{lda}.
The tangent space is obtained
To enhance classifications of \ac{erps} which rely on time domain information
which would otherwise be lost in the spatial representation of the extracted
feature, the class averages can be concatenated as additional channels to the
epochs, embedding the temporal structure in the feature~\cite{Barachant2014}.
To improve performance even further, the \acp{erp} can be decomposed by a method
like XDAWN before calculating the extended covariances~\cite{Li2020}.

@@

\todo{wrap-up}

>>>>

\todo{comments liesel paradigms}
\todo{spellcheck}
\todo{are all acronyms defined}
\todo{fix all latex warnings to make sure no refs are missing (or grep on that
specific warning)}
\todo{uniform ref or autoref}
\todo{don't capitalize figure, table, algorithm,...}
\todo{define short captions for tables and figures}
\todo{fix lacheck/chktex notifications}
\todo{use terminology individuals with severe
speech and physical impairments (SSPI), avoid phrasing 'suffering from'}
\todohsm{more subtitles}
\todofc{double check statements about what has been published 'for the sake of clarity, the introduction has been mofed to"}
\todofc{chapters should look more like full papers}
\todo{uniform vsa condition vsa setting}
\todo{acronyms written in full every new chapter and in captions}
\todo{search for ??}
\todo{acro usage in chapters}

@@

\subsection{Active \& passive \acsp{bci}}

@@

\subsection{Reactive \acsp{bci}}
Reactive \ac{bci} take another approach, by providing a discrete set of
sensory stimulations towards which the user's attention can directed.
Compared to active \ac{bci}, reactive paradigms can more easily induce fatigue
due to the constant stimulation.
These paradigms are also somewhat less intuitive compared to e.g.\ speech, and
their expressive power is limited by how many different stimuli can effectively
presented and attended within a reasonable timeframe.
Nevertheless, reactive \ac{bci} work well with \ac{eeg} recordings, and, most
importantly, they work for most people~\cite{Allison2010a,Edlinger2014}.

@@

The epochs contain values representing the voltage of a specific channel relative to the
reference at a specific time relative to the stimulus.
In \ac{erp} classification, they are usually directly used as features, by
concatenating all channels in the matrix format epoch $\mat{X}$ into one feature
vector s.t.\ $\mat{x} = \mat\vec(\mat{X})$.

\subsection{Decoders}
\todo{averaging over trials}
Several techniques are suited to decode~\acp{erp}.
Below are some methods that can achieve state-of-the art performance and are
relevant to this work.

\subsubsection{\Acf{lcmv} beamforming}
Linear decoding techniques such as the \ac{lcmv}
beamformer~\cite{Wittevrongel2016} generally work well for
classifying \ac{erp} epochs.
\Ac{lcmv} beamforming learns a weight vector $\mat{w}$ from a set of epochs that can
be multiplied with the feature vector of a new epoch to obtain a prediction
score.
The weights $\mat{w}$ are obtained by minimizing the variance of the output
\begin{equation}
  \argmin_\mat{w}\mat{w}^\intercal\mat{C}
	\mat{w}^\intercal
\end{equation}
under the linear constraint
\begin{equation}
	\mat{a}^\intercal\mat{w} = 1
\end{equation}
with activation pattern $a$ the difference between the average of the attended epochs and the
average of the non-attended epochs in the training data.
This is then solved by  method of Lagrange multipliers as
\begin{equation}
	\mat{w} =
  \frac{\mat{C}^{-1}\mat{a}^\intercal}
  {\mat{a}\mat{C}^{-1}\mat{a}^\intercal}
\end{equation}
with $\mat{C}$ the background covariance.

\subsubsection{\Acf{lda}}
\Ac{lda} is a more well-known method that is, int fact, nearly equivalent to
\ac{lcmv} beamforming~\cite{Treder2016} under certain conditions.
It assumes normally distributed data with a different mean per class but a
common covariance.
Similar to \ac{lcmv}-beamforming, \ac{lda} learns a weight vector $\mat{w}$
that satisfies
\begin{equation}
  \argmax_\mat{w}\frac{\mat{w}^\intercal\mat{S}_\text{b}\mat{w}}{\mat{w}^\intercal\mat{S}_\text{w}\mat{w}}
\end{equation}
The between-class scatter matrix $\mat{S}_\text{b}$ models the variability
between the expected responses, the within-class scatter matrix
$\mat{S}_\text{w}$ the expected noise.
The optimization criterion corresponds to finding a projection that maximizes
the spread between classes (signal) while minimizing the spread within classes
(noise).
$mat{w}$ can be obtained as eigenvector of
$\mat{S}_\text{w}^{-1}\mat{S}_\text{b}$ corresponding to the largest eigenvalue.

@@

\subsubsection{\Acf{tlda}}

Another example of this is \ac{tlda}~\cite{Sosulski2022}, which assumes temporal
stationarity on the background noise to regularize the problem.
It does this by imposing a block-Toeplitz structure to the between-class covariance
matrix of \ac{lda}, such that every temporal covariance block corresponding to
a channel has the structure
\begin{equation}
  \mat{S}_\text{w}^(c) = \begin{bmatrix}
s_0 & s_1 & s_2 & \cdots & s_{k-1} \\
s_1 & s_0 & s_1 & \cdots & s_{k-2} \\
s_2 & s_1 & s_0 & \cdots & s_{k-3} \\
\vdots & \vdots & \vdots & \ddots & \vdots \\
s_{k-1} & s_{k-2} & s_{k-3} & \cdots & s_0
\end{bmatrix}
\end{equation}
This method is simple yet effective reaching state-of-the-art decoding
performance.

@@

With this distance defined, the epochs can than be either directly be
classified on the Riemannian manifold using a minimum distance to mean
classifier, or they can be projected to the tangent space.
This is a euclidean space that locally approximates the Riemannian distance,
which allows us to apply regular machine learning methods like an SVM or
\ac{lda}.
The tangent space is obtained\todo{tangent space}.
To enhance classifications of \acp{erp} which rely on time domain information
which would otherwise be lost in the spatial representation of the extracted
feature, the class averages can be concatenated as additional channels to the
epochs, embedding the temporal structure in the feature~\cite{Barachant2014}.
To improve performance even further, the \acp{erp} can be decomposed by a method
like XDAWN before calculating the extended covariances~\cite{Li2020}.

@@

\section{The gaze-dependence problem}
